---
title: "Using RMarkdown"
teaching: 10
exercises: 2
questions:
- "How to write a lesson using RMarkdown?"
objectives:
- "Explain how to use RMarkdown with the new lesson template."
- "Demonstrate how to include pieces of code, figures, and challenges."
keypoints:
- "It shouldn't be difficult"
---
This episode demonstrates all the features that can be used when writing a
lesson in [RMarkdown][r-markdown].

This first chunk is really important, and should be included in all markdown lessons.

```{r, echo=FALSE}
source("../bin/chunk-options.R")
```

The rest of the lesson should be written as a normal RMarkdown file. You can
include chunk for codes, just like you'd normally do:

Normal output:

```{r}
1 + 1
```

Output with error message:

```{r}
x[10]
```

Output generating figures:

```{r plot-example}
library(ggplot2)
ggplot(diamonds, aes(x = carat,  y = price, color = cut)) +
    geom_point()
```

For the challenges and their solutions, you need to pay attention to the where
the `>` go and where to leave blank lines. Otherwise, you can include chunks in
it to include instructions and solutions.

> ## Challenge: Can you do it?
>
> What is the output of this command?
>
> ```{r, eval=FALSE}
> paste("This", "new", "template", "looks", "good")
> ```
>
> > ## Solution
> >
> > ```{r, echo=FALSE}
> > paste("This", "new", "template", "looks", "good")
> > ```
> {: .solution}
{: .challenge}

{% include links.md %}
