---
title: Connecting Securely to an HPC Cluster
teaching: 35
exercises: 0
---

```{r, echo=FALSE}
# Source the external configuration script
source("load_config.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Explain why secure remote access is required for HPC systems.
- Configure SSH key-based authentication for an HPC cluster.
- Connect securely to the login node of an HPC cluster.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- Why is SSH used to access HPC systems securely?
- How do I authenticate and connect to an HPC login node?

::::::::::::::::::::::::::::::::::::::::::::::::::

An HPC cluster is a collection of computers that users access remotely. Your laptop
is not part of the cluster; it is the computer from which you connect. To begin
working on the HPC cluster, you must first establish a secure connection to the
system. Because HPC clusters are shared resources accessed over institutional or
public networks, authentication and encrypted communication are fundamental
components of every HPC workflow.

## Secure Connections

The first step in using an HPC cluster is to establish a secure connection from your
local computer. When we interact with a computer, we often use *graphical user interface*,
or GUI, with windows, icons, menus, and other visual elements.
Because HPC systems are administered remotely, the *command-line interface* (CLI)
remains the standard method for interacting with HPC environments. CLI-based workflows
are lightweight, reproducible, and well suited to automation, allowing users to manage
files, compile software, execute applications, and administer remote systems efficiently.
Compared with graphical desktop environments, the CLI also consumes very little network
bandwidth.

If you have ever opened the Windows Command Prompt or macOS Terminal, you have
seen a CLI. If you have already taken "The Carpentries" courses on the UNIX
Shell or Version Control, you have used the CLI on your *local machine*
extensively. The main new concept is opening a CLI session on the remote HPC
cluster while taking appropriate precautions so that other people on the network
cannot see (or modify) the commands you run or the results returned by the remote
HPC system. We will use the _**S**ecure **Sh**ell_ protocol (or SSH) to open an
encrypted network connection between two machines, allowing commands, files, and
other data to be exchanged securely.

![connect-to-remote.svg](fig/connect-to-remote.svg){
  max-width="50%"
  alt="Connect to cluster. "
}

SSH clients are usually command-line tools that require only the remote system address
as a mandatory argument. If your username on the remote system differs from
what you use locally, you must provide that as well. If your SSH client has a
graphical front-end, such as PuTTY or MobaXterm, you will set these arguments
before clicking "connect". From the terminal, you'll write something like
`ssh username@hostname`, where the argument is similar to an email address:
the `@` separates the remote username from the remote host.
In this lesson, *`r config$remote$user`* represents the username on the remote
system, and *`r config$remote$login`* represents the cluster's login address
(`hostname`), which identifies the cluster's login node on the network.

When logging in to a laptop, tablet, or other personal device, a username, password,
or pattern are normally required to prevent unauthorized access. In these situations,
the likelihood of somebody else intercepting your password is low, since logging your
keystrokes requires a malicious exploit or physical access. For systems like
``r config$remote$host`` running an SSH server, anyone on the network can attempt
to log in to the system. Since usernames are often public or easy to guess, your
password is often the weakest link in the security chain. Many HPC centres disable
password authentication for SSH and instead require or prefer public-key authentication.
Authentication policies may vary between HPC centres, and some systems also
use institutional identity services or multi-factor authentication. The private key
should also be protected by a strong passphrase. Even if your cluster does not require
it, the next section will guide you through the use of SSH keys and an SSH agent to
both strengthen your security and make it more convenient to log in to remote systems.

### Better Security With SSH Keys

The [Lesson Setup](../learners/setup.md) provides instructions for installing
a shell application with SSH. If you have not done so already, please open that
shell application with a Unix-like command line interface to your system.

SSH authenticates users using public-key cryptography. A public key is
registered with the HPC centre's authentication system, while the corresponding
private key remains securely stored on your local computer. SSH keys provide an
alternative authentication mechanism for accessing remote computing systems.
They can also be used for authentication when transferring files or for accessing
remote version control systems (such as [GitHub][gh-ssh]). In this section you
will create a pair of SSH keys:

- a private key which you keep on your own computer, and
- a public key which can be placed on any remote system you will access.

SSH public-key authentication can replace password-based SSH authentication during login.
Instead of proving your identity with your account password, SSH proves that you possess
the private key corresponding to the public key stored on the cluster. However, your private
key should itself be protected by a passphrase. The passphrase protects your private key
while it is stored on your computer. After you enter the passphrase, the private-key
identity can be loaded into the SSH agent. The agent can then use that identity for
subsequent SSH authentication.

An SSH agent is optional. It is a convenience mechanism that holds a private-key identity
for authentication, so you do not need to enter the key's passphrase for every SSH connection.

:::::::::::::::::::::::::::::::::::::::::  caution

## Private keys are your secure digital passport

If someone else obtains your private key, treat it as compromised. Stop using it,
remove its corresponding public key from systems where it is authorized, and generate
a new key pair. This includes cases where the key (or a copy) is stored in a directory
with improper permissions, is transmitted over an unencrypted network, is sent as an
attachment via unencrypted email, or is even displayed on your terminal window.

Protect this key as if it unlocks your front door. In many ways, it does.

::::::::::::::::::::::::::::::::::::::::::::::::::

Regardless of the software or operating system you use, please choose a strong
passphrase to act as another layer of protection for your private SSH key.

:::::::::::::::::::::::::::::::::::::::::  callout

## Considerations for SSH Key Passphrases

When prompted, enter a strong passphrase that you will remember. There are two
common approaches to this:

1. Create a memorable passphrase with some punctuation and number-for-letter
  substitutions, 32 characters or longer. Street addresses work well; just
  be careful of social engineering or public records attacks.
2. Use a password manager and its built-in password generator with all
  character classes, 25 characters or longer. [KeePass][keepass] and
  [BitWarden][bitwarden] are two good options.
3. Nothing is less secure than a private key with no passphrase. If you
  skip passphrase entry by accident, go back and generate a new key pair
  with a strong passphrase.
::::::::::::::::::::::::::::::::::::::::::::::::::

#### SSH Keys on Linux, Mac, MobaXterm, and Windows Subsystem for Linux

Once you have opened a terminal, check for existing SSH keys before generating a
new key pair, since files with the same name may be overwritten.

```bash
`r config$local$prompt` ls ~/.ssh/
```

If `~/.ssh/id_ed25519` already exists, you will need to specify
a different name for the new key pair.

Generate a new public-private key pair using the following command, which will
produce a stronger key than the `ssh-keygen` default by invoking these flags:

- `-a` (default is 16): number of rounds of passphrase derivation; increase to
  slow down brute force attacks.
- `-t` (default is Ed25519): specifies the key "type" or cryptographic
  algorithm. `ed25519` specifies the [EdDSA][wiki-dsa] public-key algorithm.
- `-f` (default is /home/user/.ssh/id\_algorithm): filename to store your
  private key. The public key filename will be identical, with a `.pub`
  extension added.

```bash
`r config$local$prompt` ssh-keygen -a 100 -f ~/.ssh/id_ed25519 -t ed25519
```

When prompted, enter a strong passphrase with the
[above considerations in mind](#considerations-for-ssh-key-passphrases).
Note that the terminal will not appear to change while you type the passphrase:
this is deliberate, for your security. You will be prompted to type it again,
so don't worry too much about typos.

Take a look in `~/.ssh` (use `ls ~/.ssh`). You should see two new files:

- your private key (`~/.ssh/id_ed25519`): *do not share with anyone!*
- the shareable public key (`~/.ssh/id_ed25519.pub`): if a system administrator
  asks for a key, this is the one to send. It is also safe to upload to
  websites such as GitHub: it is meant to be seen.

:::::::::::::::::::::::::::::::::::::::::  callout

## If `ed25519` is not Supported on Your Local Computer (Use RSA for Older Systems)

If key generation failed because ed25519 is not available, try using the older
(but still strong and trustworthy) [RSA][wiki-rsa] cryptosystem. Again, first
check for an existing key:

```bash
`r config$local$prompt` ls ~/.ssh/
```

If `~/.ssh/id_rsa` already exists, you will need to specify a different filename
for the new key pair. Generate it as above, with the following extra flags:

- `-b` sets the number of bits in the key. For RSA keys, the default is 3072 bits.
  Ed25519 uses a fixed key size, so the `-b` option does not apply to Ed25519 keys.
- `-o` (no default): use the OpenSSH key format,
  rather than PEM.

```bash
`r config$local$prompt` ssh-keygen -a 100 -b 4096 -f ~/.ssh/id_rsa -o -t rsa
```

When prompted, enter a strong passphrase with the
[above considerations in mind](#considerations-for-ssh-key-passphrases).

Take a look in `~/.ssh` (use `ls ~/.ssh`). You should see two new files:

- your private key (`~/.ssh/id_rsa`): *do not share with anyone!*
- the shareable public key (`~/.ssh/id_rsa.pub`): if a system administrator
  asks for a key, this is the one to send. It is also safe to upload to
  websites such as GitHub: it is meant to be seen.
  

::::::::::::::::::::::::::::::::::::::::::::::::::

#### SSH Keys on PuTTY

If you are using PuTTY on Windows, download and use `puttygen` to generate the
key pair. See the [PuTTY documentation][putty-gen] for details.

- Select `ed25519` as the key type.
- If PuTTYgen asks for a key size, use `255` bits.
- Click on the "Generate" button.
- You do not need to enter a comment.
- When prompted, enter a strong passphrase with the
  [above considerations in mind](#considerations-for-ssh-key-passphrases).
- Save the keys in a folder no other users of the system can read.

Take a look in the folder you specified. You should see two new files:

- your private key (`id_ed25519`): *do not share with anyone!*
- the shareable public key (`id_ed25519.pub`): if a system administrator
  asks for a key, this is the one to send. It is also safe to upload to
  websites such as GitHub: it is meant to be seen.

### SSH Agent for Easier Key Handling

A private SSH key is only as secure as the passphrase used to protect or unlock it.
Entering the passphrase every time you establish an SSH connection can become tedious.
This is where the [SSH agent][ssh-agent] comes in.

Using an SSH agent, you can enter your private key's passphrase once. The agent keeps
the loaded private-key identity available for subsequent SSH connections until the
identity lifetime expires or the identity is removed from the agent. This removes the
need to enter the passphrase for every new SSH connection.

::: caution

### SSH Passphrase

Just remember your passphrase, because once it expires in the agent, you have to
type it in again.

:::

#### SSH Agents on Linux, macOS, and Windows

Open your terminal application and check if an agent is running:

```bash
`r config$local$prompt` ssh-add -l
```

- If you get an error like this one,
  
  ```error
  Error connecting to agent: No such file or directory
  ```
  
  ... then you need to launch the agent as follows:
  
  ```bash
  `r config$local$prompt` eval $(ssh-agent)
  ```
  
  :::::::::::::::::::::::::::::::::::::::::  callout
  
  ## What's in a `$(...)`?
  
  The syntax of this SSH agent command is unusual, based on what we've seen
  in the UNIX Shell lesson. This is because the `ssh-agent` command starts
  an authentication agent and prints a series of shell commands that allow
  your current shell to communicate with it -- but *does not execute them*
  automatically. 
  
  ```bash
  `r config$local$prompt` ssh-agent
  ```
  
  ```output
  SSH_AUTH_SOCK=/tmp/ssh-Zvvga2Y8kQZN/agent.131521;
  export SSH_AUTH_SOCK;
  SSH_AGENT_PID=131522;
  export SSH_AGENT_PID;
  echo Agent pid 131522;
  ```
  
  The `eval` command interprets this text output as commands and allows you
  to access the SSH agent connection you just created.
  
  You could run each line of the `ssh-agent` output yourself, and
  achieve the same result. Using `eval` just makes this easier.
  
  
  ::::::::::::::::::::::::::::::::::::::::::::::::::

- Otherwise, your agent is already running: don't mess with it.

Add your key to the agent, with session expiration after 8 hours:

```bash
`r config$local$prompt` ssh-add -t 8h ~/.ssh/id_ed25519
```

```output
Enter passphrase for .ssh/id_ed25519: 
Identity added: .ssh/id_ed25519
Lifetime set to 28800 seconds
```

For the next eight hours, the SSH agent can use the loaded private-key identity whenever
it is needed, so you don't have to enter its passphrase again.

#### SSH Agent on PuTTY

If you are using PuTTY on Windows, download and use `pageant` as the SSH agent.
See the [PuTTY documentation][putty-agent].

### Uploading Your Public Key

```{r, eval=!is.null(config$remote$portal) && nzchar(trimws(config$remote$portal)), echo=FALSE, results="asis"}
cat(
    paste(
        "Visit [",
        config$remote$portal,
        "](",
        config$remote$portal,
        ") to upload your SSH public key. (Remember, it's the one ending in `.pub`!)",
        sep="")
)
```

```{r, eval=is.null(config$remote$portal) || !nzchar(trimws(config$remote$portal)), echo=FALSE, results="asis"}
cat(
    paste(
        "No SSH key upload portal is configured for this cluster.",
        "If supported, `ssh-copy-id` or `scp` can be used for registering or transferring the public key."
        )
)
```

## Log In to the Cluster

Go ahead and open your terminal or graphical SSH client, then log in to the
cluster. Replace ``r config$remote$user`` with your username or the one
supplied by the instructors.

```bash
`r config$local$prompt` ssh `r config$remote$user`@`r config$remote$login`
```

Depending on your cluster configuration, SSH may prompt you for either your account
password or your private key's passphrase. Watch out: characters typed at either
prompt are not displayed on the screen. Normal output will resume once you press `Enter`.

You may have noticed that the prompt changed when you logged into the remote
system using the terminal (if you logged in using PuTTY this will not apply
because it does not offer a local terminal). This change is important because
it can help you distinguish on which system the commands you type will be run
when you pass them into the terminal. This change is also a small complication
that we will need to navigate throughout the workshop. Exactly what is displayed
as the prompt (which conventionally ends in `$`) in the terminal when it is
connected to the local system and the remote system will typically be different
for every user. We still need to indicate which system we are entering commands
on though so we will adopt the following convention:

- ``r config$local$prompt`` when the command is to be entered on a terminal
  connected to your local computer
- ``r config$remote$prompt`` when the command is to be entered on a
  terminal connected to the remote system
- `$` when it really doesn't matter which system the terminal is connected to.

## Looking Around Your Remote Home

Very often, many users are tempted to think of a high-performance computing
installation as one giant, magical machine. Sometimes, people will assume that
the computer they've logged onto is the entire computing cluster. So what's
really happening? What computer have we logged on to? The name of the current
computer we are logged onto can be checked with the `hostname` command. (You
may also notice that the current hostname is also part of our prompt!)

```bash
`r config$remote$prompt` hostname
```

```output
`r config$remote$host`
```

The `hostname` identifies the computer to which you are currently connected. In most HPC systems,
this computer is known as the *login node*. The login node provides an interactive environment
where users authenticate, manage files, compile applications (where permitted by local policy),
and prepare computational work. Later in the course, you will learn how different types of jobs are
submitted from the login node for execution.

::: callout

## Login Nodes

A login node is typically one of the primary entry points to an HPC cluster. Use it to:

- authenticate;
- inspect and manage files;
- configure your environment;
- prepare your work;
- submit jobs;
- monitor jobs.

It is not recommended to run long-running or resource-intensive scientific workloads directly
on a login node unless the cluster documentation explicitly permits it.

:::


Next, let's find out where we are by running `pwd` to **p**rint the **w**orking **d**irectory.

```bash
`r config$remote$prompt` pwd
```

```output
`r paste(config$remote$fs$home,config$remote$user,sep="/")`
```

The output shows your current working directory on the HPC system. By default, your SSH session
begins in your home directory, where your personal files, shell configuration files, and other
user-specific configurations may be stored.

```bash
`r config$remote$prompt` ls
```

```output

```

The system administrators may have configured your home directory with some helpful
files, directories (or folders), and symbolic links (shortcuts) to space
reserved for you on other filesystems. If they did not, your home directory may appear
empty. To double-check, include hidden files in your directory listing:

```bash
`r config$remote$prompt` ls -a
```

```output
  .            .bashrc
  ..           .ssh
```

In the first column, `.` is a reference to the current directory and `..` a
reference to its parent (``r config$remote$fs$home``). You may or may not see the
other files, or files like them: `.bashrc` is a shell configuration file, which
you can edit with your preferences; and `.ssh` is a directory containing user-specific
SSH configurations and authentication-related files. Its contents differ between
your local computer and your HPC account.

### Verify Your SSH Key Registration

:::::::::::::::::::::::::::::::::::::::::  callout

## SSH Key Registration Depends on the HPC Centre

Policies and practices for handling SSH keys vary between HPC clusters:
follow any guidance provided by the cluster administrators or
documentation. In particular, if there is an online portal for managing SSH
keys, use that instead of the directions outlined here.

::::::::::::::::::::::::::::::::::::::::::::::::::

If you uploaded your SSH public key through the [cluster's web portal](#uploading-your-public-key),
the key has been registered with the authentication system used by the cluster. On systems that
use a conventional `~/.ssh/authorized_keys` file, you may see the registered key there.

```bash
`r config$remote$prompt` ls ~/.ssh
```

```output
authorized_keys
```

::: caution

HPC centres may store and manage SSH public keys in different ways. Your centre
may use a web portal, a local `authorized_keys` file, or another authentication
service. Do not be surprised if your HPC account does not contain a `~/.ssh` directory
or an `authorized_keys` file.

:::

The contents of `~/.ssh` vary between systems and can contain additional files such as
`known_hosts` or `config`.

If your HPC centre uses `~/.ssh/authorized_keys`, you can inspect the file with:

```bash
`r config$remote$prompt` cat ~/.ssh/authorized_keys
```

Each line in `authorized_keys` contains a public key that the SSH server may accept
for public-key authentication to this account. If you access the cluster from multiple
computers, this file may contain multiple entries.

```output
ssh-ed25519 AAAAC3NzaC1lZDI1NTE5AAAAIJC/bo0gZHk+8/kl8DNFd8hUrUqaGdL7o+TisQe/uQlp you@laptop
```

Note that the output shown above can vary from person to person and location to location.

If your HPC centre does not provide a portal for registering SSH keys, follow the key-registration
method documented by the centre. On systems that permit direct shell-based registration,
`ssh-copy-id` or an `scp`-based workflow may be available. These are alternative bootstrap
workflows for HPC centres that permit users to register SSH keys directly from the command
line. Use them only when the centre's documentation permits this workflow.

::: spoiler

## Alternative: Registering a Key with `ssh-copy-id`

`ssh-copy-id` is a bootstrap mechanism: it normally uses an authentication method
that is already available to install the public key for future public-key authentication.

`ssh-copy-id` uses `ssh` to connect to the remote system and, by default, attempts
to authenticate using the available authentication mechanisms. In the usual case,
this means using the remote account's login password, so password authentication
must be enabled on the SSH server.

The command examines the keys available to the local SSH client and determines
which keys are already authorized on the remote system. When an SSH agent is available,
its keys can be used during this process.

With the `-i ~/.ssh/id_ed25519.pub` option, we explicitly select the public key that
should be installed. This prevents `ssh-copy-id` from selecting other identities
that may be available through the SSH agent or default identity files.

`ssh-copy-id` normally installs the selected public key in the remote account's
authorized-key configuration, commonly `~/.ssh/authorized_keys`.
If the `~/.ssh` directory or `authorized_keys` file does not yet exist, `ssh-copy-id`
can create them as necessary.

```bash
`r config$local$prompt` ssh-copy-id -i ~/.ssh/id_ed25519.pub `r config$remote$user`@`r config$remote$login`
```

Some SSH servers require restrictive permissions on the `.ssh` directory and
the `authorized_keys` file before key-based authentication is accepted.

```bash
`r config$remote$prompt` chmod 700 ~/.ssh
`r config$remote$prompt` chmod 640 ~/.ssh/authorized_keys
```

::: callout

Depending on the SSH server configuration, overly permissive ownership or permissions
on these files can cause the server to reject public-key authentication.

:::

::::

::: spoiler

## Alternative: Transfer a Key with `scp`

If the `authorized_keys` file does not exist:
You need to "register" the public-key: it must be listed in a file named `authorized_keys`,
under the `.ssh` folder.

If the `.ssh` folder was not listed on the remote cluster, then it does not yet exist: create it.

```bash
`r config$remote$prompt` mkdir ~/.ssh
```

From your local computer, transfer only the public key, whose filename ends in `.pub`,
to your HPC home directory.

```{r, echo=FALSE, results="asis"}
cat(
    paste(
        "```bash\n",
        config$local$prompt,
        " scp ~/.ssh/id_ed25519.pub ",
        config$remote$user,
        "@",
        config$remote$login,
        ":~/\n```",
        sep="")
)
```

List the hidden files in the directory:

```bash
`r config$remote$prompt` ls -a
```

```output
  .            .bashrc           id_ed25519.pub
  ..           .ssh
```

Now, use `cat` to print your public key, but redirect the output, appending it
to the `authorized_keys` file:

```bash
`r config$remote$prompt` cat ~/id_ed25519.pub >> ~/.ssh/authorized_keys
```

Some SSH servers require restrictive permissions on the `.ssh` directory and
the `authorized_keys` file before key-based authentication is accepted.

```bash
`r config$remote$prompt` chmod 700 ~/.ssh
`r config$remote$prompt` chmod 640 ~/.ssh/authorized_keys
```

::: callout

Depending on the SSH server configuration, overly permissive ownership or permissions
on these files can cause the server to reject public-key authentication.

:::

::::

To verify that key-based authentication works for a new SSH session, disconnect from
the cluster and connect again:

```bash
`r config$remote$prompt` logout
```

If your private key is loaded in the SSH agent, you normally will not need to enter
its passphrase again while that identity remains available to the agent.

```bash
`r config$local$prompt` ssh `r config$remote$user`@`r config$remote$login`
```

[gh-ssh]: https://docs.github.com/en/authentication/connecting-to-github-with-ssh
[keepass]: https://keepass.info
[bitwarden]: https://bitwarden.com
[wiki-rsa]: https://en.wikipedia.org/wiki/RSA_\(cryptosystem\)
[wiki-dsa]: https://en.wikipedia.org/wiki/EdDSA
[putty-gen]: https://tartarus.org/~simon/putty-prerel-snapshots/htmldoc/Chapter8.html#pubkey-puttygen
[ssh-agent]: https://www.ssh.com/academy/ssh/agent
[putty-agent]: https://tartarus.org/~simon/putty-prerel-snapshots/htmldoc/Chapter9.html#pageant

:::::::::::::::::::::::::::::::::::::::: keypoints

- HPC clusters are shared computational resources that users access remotely.
- Secure Shell (SSH) provides encrypted communication between your local computer
  and an HPC cluster.
- SSH public-private key pairs provide a secure method for authenticating to
  remote systems.
- A private key remains on your local computer and should be protected with a strong
  passphrase.
- An SSH agent can temporarily retain a private-key identity so that its passphrase
  does not need to be entered on every connection.
- Users typically connect to an HPC system through a login node, where they authenticate,
  manage files, configure their environment, and prepare their work.

::::::::::::::::::::::::::::::::::::::::::::::::::
