---
title: Parallelising with Job Arrays
teaching: 15
exercises: 5
---

```{r, echo=FALSE}
# Source the external configuration script
source("load_config.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Prepare a job submission script for an array job.
- Launch a job to be executed in parallel over several nodes

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are job arrays?
- What benefits do job arrays bring?
- What type of jobs would benefit from job arrays?

::::::::::::::::::::::::::::::::::::::::::::::::::

Parallel computing is a technique used to divide big tasks into smaller ones
that can be solved simultaneously. Parallelism can be accomplished in different
ways and it depends on the tasks that needs doing as well as the algorithms
implemented to perform these tasks.

One way of implementing parallel computing is to distribute a job across
multiple processors. This is usually accomplished by using the Message Passage
Interface (MPI) which is a standardised way for CPU cores to communicate with
one another while working together on a task. Software has to be written
specifically to utilize MPI to take advantage of this. We will see more
of this in a later episode.

Another, simpler form of parallel computing is an array job. This type of
job is advantageous if the same software has to be run several times
in a way where the separate runs do not share data, and do not need
to run concurrently. For instance,
you might need to run the same task on several independent input
files, or you may have multiple serial tasks that take some parameter,
and you need to explore several values of the parameter. These
tasks are also sometimes called "high-throughput" computing.

:::::::::::::::::::::::::::::::::::::::  challenge

What is the important distinguishing feature between tasks
that are appropriate for array jobs and task that are not?
Name some examples.

:::::::::::::::  solution

Tasks appropriate for array jobs are "high-throughput", where
the same thing needs to be done many times, possibly over a set
of parameters, but where each task in dependent of the others.

For example, running the same statistical analysis on a large
number of independent input files is a good candidate for an
array solution.

Parallel tasks which have interactions between the various
parallel threads need to communicate between tasks at
run-time, and are not appropriate for job arrays.

For example, most parallel scientific codes that run in
parallel have a requirement to communicate between parallel
elements at run-time, and are not appropriate for job arrays.

Similarly, serial tasks which only need to be run once do not
benefit from parallelism. Aggregating unrelated tasks into an
array merely for the sake of grouping does not make sense.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


We have aleady seen how to submit a single serial job to the
Slurm scheduler, and also a single parallel jobs.

Array jobs differ in character, but are specified in a similar manner.
The batch file describes the array to the system, and at submission
time, all of the array elements are submitted together. They may
dispatch at different times, and may or may not run concurrently.

Within
the execution envioronment of an array element, a number of environment
variables will be set, which can be used at run-time by the
array elements to identify the size of the array, and their
position within it, allowing them to select the correct input
file. These jobs will be dispatched as their resource requirements
and priorities are met, and they may or may not run concurrently.

By way of review, we will construct a job submission script that
runs the `amdahl` executable with the default arguments.

Using `nano`, create a script called
`job_single_amdahl.sh` containing the following:

```{r, echo=FALSE, results='asis'}
snippets('array/job_single_amdahl.Rmd')
```

:::::::::::::::::::::::::::::::::::::::  challenge

How would you submit the script to Slurm for execution?

:::::::::::::::  solution

  
```bash
sbatch job_single_amdahl.sh
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Array Job Syntax


To specify an array job, you only need to add a single
directive to your batch file, and then adapt your run command
to take advantage of the information provided by the environment
variables.

The relevant array directive has this format:

```bash
`r config$sched$comment` `r config$sched$flag$array`=<array-spec>
```

The `<array-spec>` above is a place-holder for specifying the
size and extent of the array. The specification will resolve to
set of integers, which will index the job array elements.

For a simple example, an array specification of `1-4` means the
system should create four array elements, numbered consecutively
from one through four.

You can also specify a comma-separated set of numbers, such
as `1,3,5`, or you can specify a stride, for example by specifying
`1-10:2` (which is equivalent to `1,3,5,7,9`).

In addition to these, you can also specify a limit on the number
of array elements that will run concurrently, using the `%` sign.
An example of this, building on what we saw before, would be
to specify `1-10:2%4`, which will create five array elements
with indices 1, 3, 5, 7, and 9, and run at most four of them
at a time until they are all complete.

When an array element job is running, the run-time environment
will include some special environment variables, the most important
of which is `SLURM_ARRAY_TASK_ID`, which specifies the index
of the current instance.  There are other environment variables
which tell you the full size of the array, and the starting and
ending indices. As we have seen, because there is a fairly rich
syntax for specifying arrays, it is not straightforward to
infer the size of the array from the high and low indices.

There are also some file-name patterns you can use to control
where your executable reads and writes data. The most important
of these is the `%a` pattern, which corresponds to the index
of the current array element, similarly to `SLURM_ARRAY_TASK_ID`.

::::: challenge


Write a batch script to run the `amdahl` program as an array job with
4 independent jobs, to examine the effects of jitter.

Use the directives discussed above to specify the array elements,
and to direct the output of each array element to a separate file.

:::: solution

```{r, echo=FALSE, results='asis'}
snippets('array/job_array_amdahl.Rmd')
```

::::

:::::


:::: instructor

You can download the script from https://raw.githubusercontent.com/NewcastleRSE-Training/hpc-intro-comet/refs/heads/main/episodes/files/job_array_word-freq.sh

::::


:::::::::::::::::::::::::::::::::::::::: keypoints

- Parallel programming allows applications to take advantage of parallel hardware.
- The queuing system facilitates executing parallel tasks.
- Parallel computing allows applications to distribute the workload over several CPUs or nodes
- There are multiple parallelization strategies that are generally supported by resource managers.
- Array parallel jobs are suitable for independent runs of the same executable with varying inputs or outputs. 

::::::::::::::::::::::::::::::::::::::::::::::::::
