---
title: Parallelising with Job Arrays
teaching: 15
exercises: 5
---

```{r, echo=FALSE}
# Source the external configuration script
source("load_config.R")
```

::::::::::::::::::::::::::::::::::::::: objectives

- Prepare a job submission script for an array job.
- Launch a job to be executed in parallel over several nodes

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are job arrays?
- What benefit does job arrays bring?
- What type of jobs would benefit from job arrays?

::::::::::::::::::::::::::::::::::::::::::::::::::


Parallel computing is a technique used to divide big tasks into smaller ones
that can be solved simultaneously. Parallelism can be accomplished in different
ways and it depends on the tasks that needs doing as well as the algorithms
implemented to perform these tasks.

One way of implementing parallel computing is to distribute a job across
multiple processors. This is usually accomplished by using the Message Passage
Interface (MPI) which is a standardised way for CPU cores to communicate with
one another while working together on a task. Software has to be written
specifically to utilize MPI to take advantage of this. We will see more
of this in a later episode.

Another, simpler form of parallel computing is an array job. This type of
job is advantageous if the same software has to be run several times
in a way where the separate runs do not share data, and do not need
to run concurrently. For instance,
you might need to run the same task on several independent input
files, or you may have multiple serial tasks that take some parameter,
and you need to explore several values of the parameter. These
tasks are also sometimes called "high-throughput" computing.

:::::::::::::::::::::::::::::::::::::::  challenge

What is the important distinguishing feature between tasks
that are appropriate for array jobs and task that are not?
Name some examples.

:::::::::::::::  solution

Tasks appropriate for array jobs are "high-throughput", where
the same thing needs to be done many times, possibly over a set
of parameters, but where each task in dependent of the others.

For example, running the same statistical analysis on a large
number of independent input files is a good candidate for an
array solution.

Parallel tasks which have interactions between the various
parallel threads need to communicate between tasks at
run-time, and are not appropriate for job arrays.

For example, most parallel scientific codes that run in
parallel have a requirement to communicate between parallel
elements at run-time, and are not appropriate for job arrays.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


We have aleady seen how to submit a single serial job to the
Slurm scheduler. Array jobs are specified in a similar manner.
You describe the full array to the scheduler through the submission
file, and submit it to the system with the same command
we used before.

The Slurm system will create all of the array elements. Within
the execution envioronment of a job, a number of environment
variables will be set, which can be used at run-time by the
array elements to identify the size of the array, and their
position within it, allowing them to select the correct input
file. These jobs will be dispatched as their resource requirements
and priorities are met, and they may or may not run concurrently.


```bash
cd training directory
mkdir username
cd username
```

:::: challenge

Download the word frequency script:

  `wget` [word-freq.sh](files/word-freq.sh)

::::

write a small file to test our script

```bash
`r config$remote$prompt` nano test-data.txt
`r config$remote$prompt` cat test-data.txt
```

```output
This is a small file - it will be very useful for trying out our script.
Some words are repeated in this file
- we can look for repeated words
and count them (to see which words are repeated most often).
```

To test the script we will run it on the login node. Remember, never do this
with a resource intensive script. You could even run the script on your laptop
or desktop if it uses Linux or Mac. This specific script will not work on
Windows as not all the commands in the script are available on the Windows
operating system.

```bash
bash word-freq.sh test-data.txt
```

To the results, type the output of the script to screen:

```bash
cat analyzed_test-data.txt
```

You should get something like this:

```{r, echo=FALSE, results='asis'}
snippets('array/analyzed_test-data.Rmd')
```

Once we have proved that the script runs without a problem we can write a
script that can be submitted to Slurm. Using `nano`, create a script called
`job_single_word-freq.sh` containing the following:

```{r, echo=FALSE, results='asis'}
snippets('array/job_single_word-freq.Rmd')
```

:::::::::::::::::::::::::::::::::::::::  challenge

How would you submit the script to Slurm for execution?

:::::::::::::::  solution

  
```bash
sbatch job_single_word-freq.sh
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

If you haven't done so already, download the data to be used for the
script. The data consists of several books from the Gutenberg project as text
files. The downloaded books should be as follows:

| File name | Author           | Book               |
|-----------|------------------|--------------------|
| data.1    | Shakespeare      | Collected Works    |
| data.2    | Geoffrey Chaucer | _Canterbury Tales_ |
| data.3    | Herman Melville  | _Moby Dick_        |
| data.4    | Homer            | _The Odyssey_      |

::::: challenge

Download the data using [the __make-data.sh__ script](files/make-data.sh).

:::::

::::: challenge

How would we change the `job_single_word-freq.sh` script to use the first of
the four data files instead of `test-data.txt`?

:::: solution

Write `data.1` in place of `test-data.txt`.
::::

:::::


## Array Jobs

TO DO:

- Introduce the concept of Slurm Job Arrays. Why might we want to use one?
- Introduce the syntax of Slurm Job Arrays. What do the hyphenated numerals represent? Where do they start?
- Keep cognitive burden as light as practicable. A lot has been burned on preparing the synthetic datasets.

::::: challenge

Write a batch script to call `word-freq.sh` as an array job with 4 parallel
jobs to process all 4 text files (job_array_word-freq.sh). To do this you will
need the directive `#SBATCH --array=1-4`. When using this directive, each job
will be given a job number. In this case it will be job numbers one to
four. While running the script for a specific job number, that number will be
available in an environment variable called `${SLURM_ARRAY_TASK_ID}`.

:::: solution

```{r, echo=FALSE, results='asis'}
snippets('array/job_array_word-freq.Rmd')
```

::::

:::::


:::: instructor

You can download the script from https://raw.githubusercontent.com/NewcastleRSE-Training/hpc-intro-comet/refs/heads/main/episodes/files/job_array_word-freq.sh

::::


:::::::::::::::::::::::::::::::::::::::: keypoints

- Parallel programming allows applications to take advantage of parallel hardware.
- The queuing system facilitates executing parallel tasks.
- Parallel computing allows applications to distribute the workload over several CPUs or nodes
- Parallelising over CPUs uses MPI (Message Passing Interface) -- FIXME: MPI is not involved with this software
- Parallelising over nodes can be accomplished using array jobs -- FIXME: conceptual non-sequitur, as the scripts only allocate one node

::::::::::::::::::::::::::::::::::::::::::::::::::
