---
title: Parallelising with Job Arrays
teaching: 15
exercises: 5
---

::::::::::::::::::::::::::::::::::::::: objectives

- Prepare a job submission script for an array job.
- Launch a job to be executed in parallel over several nodes

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What are job arrays?
- What benefits do job arrays bring?
- What type of jobs would benefit from job arrays?

::::::::::::::::::::::::::::::::::::::::::::::::::

Parallel computing allows multiple computational tasks to execute
simultaneously in order to reduce execution time for a task, or
increase throughput for multiple tasks. Depending on the application,
the workload may be divided into cooperating subtasks that communicate with
one another, or into independent tasks that execute separately.

One common approach to parallel computing is to distribute computation across
multiple processes that cooperate by exchanging information during
execution using the Message Passing Interface. This is the "MPI" that we
saw as a dependency of the `amdahl` application.

Not all workloads require processes to cooperate. Many scientific workflows
are made up of independent jobs. For these cases, `r config$sched$name`
provides Job Arrays, allowing many similar jobs to be submitted and managed together.

For instance,
you might need to run the same task on several independent input
files, or you may have multiple serial tasks that take some parameter,
and you need to explore several values of the parameter.
Workflows made up of these independent elements are 
also sometimes called "high-throughput" computing.

Operationally speaking, a Job Array is a collection of related batch
jobs submitted using a single job script.

:::::::::::::::::::::::::::::::::::::::  challenge

What distinguishes workloads that are suitable for job arrays from those
that require traditional parallel programming?
Describe some examples.

:::::::::::::::  solution

Tasks appropriate for array jobs are "high-throughput", where
the same thing needs to be done many times, possibly over a set
of parameters, but where each task is independent of the others.

For example, running the same statistical analysis on a large
number of independent input files is a good candidate for an
array solution.

Parallel tasks which have interactions between the various
parallel processes need to communicate between processes at
run-time, and are not appropriate for job arrays.

For example, most parallel scientific codes that run in
parallel have a requirement to communicate between parallel
elements at run-time, and are not appropriate for job arrays.

Similarly, serial tasks which only need to be run once do not
benefit from parallelism. Aggregating unrelated tasks into an
array merely for the sake of grouping does not make sense.

:::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::::::

We have already seen how to submit a single serial job to the
Slurm scheduler.

Array jobs differ in character, but are specified in a similar manner.
The batch file describes the array to the system, and at submission
time, all of the array elements are submitted together. They may
dispatch at different times, and may or may not run concurrently.

Within
the execution environment of an array element, a number of environment
variables will be set, which can be used at run-time by the
array elements to identify the size of the array, and their
position within it, allowing them to select the correct input
file. These jobs will be dispatched as their resource requirements
and priorities are met, and they may or may not run concurrently.

By way of review, we will construct a job submission script that
runs the `amdahl` executable with the default arguments.

Using `nano`, create a script called
`job_single_amdahl.sh` containing the following:

```{r, echo=FALSE, results='asis'}
snippets('array/job_single_amdahl.Rmd')
```

:::::::::::::::::::::::::::::::::::::::  challenge

How would you submit the script to Slurm for execution?

:::::::::::::::  solution

  
```bash
sbatch job_single_amdahl.sh
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Array Job Syntax


To specify an array job, you only need to add a single
directive to your batch file, and then adapt your run command
to take advantage of the information provided by the environment
variables.

The relevant array directive has this format:

```bash
`r config$sched$comment` `r config$sched$flag$array`=<array-spec>
```

The `<array-spec>` above is a place-holder for specifying the
size and extent of the array. The specification will resolve to
set of integers, which will index the job array elements.

For a simple example, an array specification of `1-4` means the
system should create four array elements, numbered consecutively
from one through four.

You can also specify a comma-separated set of numbers, such
as `1,3,5`, or you can specify a stride, for example by specifying
`1-10:2` (which is equivalent to `1,3,5,7,9`).

In addition to these, you can also specify a limit on the number
of array elements that will run concurrently, using the `%` sign.
An example of this, building on what we saw before, would be
to specify `1-10:2%4`, which will create five array elements
with indices 1, 3, 5, 7, and 9, and run at most four of them
at a time until they are all complete.

When an array element job is running, the run-time environment
will include some special environment variables, the most important
of which is `SLURM_ARRAY_TASK_ID`, which specifies the index
of the current instance.  There are other environment variables
which tell you the full size of the array, and the starting and
ending indices. As we have seen, because there is a fairly rich
syntax for specifying arrays, it is not straightforward to
infer the size of the array from the high and low indices.

There are also some file-name patterns you can use to control
where your executable reads and writes data. The most important
of these is the `%a` pattern, which corresponds to the index
of the current array element, similarly to `SLURM_ARRAY_TASK_ID`.

::::: challenge


Write a batch script to run the `amdahl` program as an array job with
4 independent jobs, to examine the effects of jitter.

Use the directives discussed above to specify the array elements,
and to direct the output of each array element to a separate file.

:::: solution

```{r, echo=FALSE, results='asis'}
snippets('array/job_array_amdahl.Rmd')
```

::::

:::::


:::::::::::::::::::::::::::::::::::::::: keypoints

- Parallel programming allows applications to take advantage of parallel hardware.
- The queuing system facilitates executing parallel tasks.
- Parallel computing allows applications to distribute the workload over several CPUs or nodes
- There are multiple parallelization strategies that are generally supported by resource managers.
- Array parallel jobs are suitable for independent runs of the same executable with varying inputs or outputs. 

::::::::::::::::::::::::::::::::::::::::::::::::::
